Very simplified likelihood

### 2024-08-21 inst/very_simplified_likelihood.Rmd
replace with actual data
start work on #240

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=1+b(n-1)$.

Suppose our bro comes first (zero $b$'s ahead of him):

@@

Comes second, one $b$ ahead of him:

@@

Third, two ahead of him:

@@

Fourth, three ahead:

@@

@@

@@

@@

@@

```{r}

rank_auck    <- c(177, 222, 206, 142, 118, 224, 128, 115, 183)
runners_auck <- c(259, 305, 297, 241, 179, 338, 203, 245, 254)

rank_stir <-    c(145, 157, 156, 191, 172, 174, 173, 165, 172, 199, 181, 229)
runners_stir <- c(193, 240, 197, 257, 227, 234, 238, 238, 196, 242, 242, 318)
```

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $R$, the rank of the focal competitor: we
have $1\leqslant R\leqslant n+1$.  Alternatively we might consider
$r=R-1$, the number of $b$ clones finishing ahead of $a$, so
$0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=1+b(n-1)$.

$$r=R-1$$

Suppose our bro comes first (zero $b$'s ahead of him) [$r=0, R=1$]:

@@

Comes second [$r=1,R=2$], one $b$ ahead of him:

@@

Third [$r=2, R=3$]; two ahead of him:

@@

Fourth [$r=3, R=4$], three ahead:

@@

\[
\mathcal{L}_{R,n}(a)=\frac{(B-1)(B+n-R-1)!}{(B+n-1)!},\qquad B=1/(1-a)
\]

\[
=\frac{(B-1)\Gamma(B+n-R)}{\Gamma(B+n)},\qquad B=1/(1-a)
\]

@@

\[
\log(B-1) + ((B+n-R-1)\log(B+n-R-1)-(B+n-R-1)+\frac{\log(B+n-R-1)}{2})
          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
\]

@@

Now using $R$ not $r$:

```
sage: var('B n R')
(B, n, R)
pretty_print_default(False)
X = log(B-1) + (
  + ((B+n-R-1)*log(B+n-R-1)-(B+n-R-1)+log(B+n-R-1)/2)
  - ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
  )
X
(B - R + n - 1)*log(B - R + n - 1) - (B + n - 1)*log(B + n - 1) + R + 1/2*log(B - R + n - 1) - 1/2*log(B + n - 1) + log(B - 1)
X.taylor(n,Infinity,2).full_simplify()
-1/12*(12*R*n^2*log(n) + 3*(2*B - 1)*R^2 - 2*R^3 - 12*n^2*log(B - 1) - 6*(B^2 - B)*R + 6*((2*B - 1)*R - R^2)*n)/n^2
```

@@

Now some Mathematica:

```
f[B_, n_, R_] := Log[(B - 1)*Gamma[B + n - R]/Gamma[B + n]] 
Assuming[B > 1 && p > 0 && p < 1, 
 Series[f[B, n, n*p], {n, Infinity, 1}]]
```

We get

$$
\log\left(
(B-1)(1-p)^{B-1/2} +
\frac{
	(B-1)(1-6B+6B^2)(1-p)^{B-1/2}p
}{
	12(1-p)n
}
\right)
$$

or

$$
\log\left(
(B-1)(1-p)^{B-1/2} \left[1+
\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
$$

$$
\log
(B-1) + B\log(1-p) + 
\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
$$

@@

```{r}
a <- read.table("parkrun_results.txt",header=TRUE)

rank_auck    <- a$place  [a$venue == "auckland"]
runners_auck <- a$runners[a$venue == "auckland"]

rank_stir    <- a$place  [a$venue == "stirling"]
runners_stir <- a$runners[a$venue == "stirling"]

rank_auck    
runners_auck

rank_stir 
runners_stir
